---
title: "Importing data"
output:
  html_document
---

```{r include=FALSE}
options(width=110)
knitr::opts_chunk$set(comment=NA, fig.align="center")
```

## Prerequisites
In this chapter, we will use the [readr](https://readr.tidyverse.org/index.html) package to import rectangular data stored in plain text files. So far, we have only worked with example data already available in R, but at some point we need to import our own datasets. Unfortunately, it is rarely the case that these datasets are already tidy. Instead, real-world data may contain typos and/or inconsistent notations, combine multiple variables in one column, use multiple columns for a single variable, and so on.

In addition to readr, the Tidyverse also provides packages for importing data from other statistical software. In particular, the [haven](https://haven.tidyverse.org/) package supports SPSS, Stata, and SAS formats, whereas [readxl](https://readxl.tidyverse.org/) imports data stored in Excel sheets.

We start with importing plain text files:

```{r}
library(readr)
```

## Plain text files
Plain text files contain only text (letters, numbers, punctuation, and other special characters). They can be opened by any text editor on almost any platform. Therefore, storing data in a plain text file is often a viable option unless there are of hundreds of thousands of rows and columns. Such large datasets would take up a lot of disk space when stored as plain text, so other more specialized binary formats (not covered in this workshop) should be preferred.

In many cases, data can be represented as a table consisting of rows and columns. Therefore, our goal is to convert rectangular data from its text representation to a data frame in R.

Here's the contents of *mtcars.csv*, an example rectangular plain text file that comes with readr (you can find its location on your computer with `readr_example("mtcars.csv")`):

```{}
"mpg","cyl","disp","hp","drat","wt","qsec","vs","am","gear","carb"
21,6,160,110,3.9,2.62,16.46,0,1,4,4
21,6,160,110,3.9,2.875,17.02,0,1,4,4
22.8,4,108,93,3.85,2.32,18.61,1,1,4,1
21.4,6,258,110,3.08,3.215,19.44,1,0,3,1
18.7,8,360,175,3.15,3.44,17.02,0,0,3,2
18.1,6,225,105,2.76,3.46,20.22,1,0,3,1
14.3,8,360,245,3.21,3.57,15.84,0,0,3,4
24.4,4,146.7,62,3.69,3.19,20,1,0,4,2
22.8,4,140.8,95,3.92,3.15,22.9,1,0,4,2
19.2,6,167.6,123,3.92,3.44,18.3,1,0,4,4
17.8,6,167.6,123,3.92,3.44,18.9,1,0,4,4
16.4,8,275.8,180,3.07,4.07,17.4,0,0,3,3
17.3,8,275.8,180,3.07,3.73,17.6,0,0,3,3
15.2,8,275.8,180,3.07,3.78,18,0,0,3,3
10.4,8,472,205,2.93,5.25,17.98,0,0,3,4
10.4,8,460,215,3,5.424,17.82,0,0,3,4
14.7,8,440,230,3.23,5.345,17.42,0,0,3,4
32.4,4,78.7,66,4.08,2.2,19.47,1,1,4,1
30.4,4,75.7,52,4.93,1.615,18.52,1,1,4,2
33.9,4,71.1,65,4.22,1.835,19.9,1,1,4,1
21.5,4,120.1,97,3.7,2.465,20.01,1,0,3,1
15.5,8,318,150,2.76,3.52,16.87,0,0,3,2
15.2,8,304,150,3.15,3.435,17.3,0,0,3,2
13.3,8,350,245,3.73,3.84,15.41,0,0,3,4
19.2,8,400,175,3.08,3.845,17.05,0,0,3,2
27.3,4,79,66,4.08,1.935,18.9,1,1,4,1
26,4,120.3,91,4.43,2.14,16.7,0,1,5,2
30.4,4,95.1,113,3.77,1.513,16.9,1,1,5,2
15.8,8,351,264,4.22,3.17,14.5,0,1,5,4
19.7,6,145,175,3.62,2.77,15.5,0,1,5,6
15,8,301,335,3.54,3.57,14.6,0,1,5,8
21.4,4,121,109,4.11,2.78,18.6,1,1,4,2
```


## The `readr` functions
The most generic function for reading rectangular data is called `read_delim()`. It has parameters to set the column delimiter, decimal mark, missing values, whether or not there is a header row, and so on. Other functions include `read_csv()`, `read_tsv()`, and `read_csv2()`, which are special cases with predefined values for some common file formats. We'll showcase how readr functions work using `read_delim()`, because all functions behave very similarly.

The first argument is always the file name you want to import. The second argument specifies the column separator used in the particular file. For example, if you wanted to load *mtcars.csv* shown in the previous example, you could use this function call to store the contents of the file in a data frame `df`:

```{r message=FALSE}
df <- read_delim(readr_example("mtcars.csv"), ",")
```

However, data files are usually located in the current working directory. This allows us to specify the file name without the complete path as follows:

```{r eval=FALSE}
df <- read_delim("mtcars.csv", ",")
```


## Comparison to base R functions


## Parsing a vector

## Parsing a file

## Writing to a file

## Example datasets
### COVID19 in Austria
https://covid19-dashboard.ages.at/data/CovidFaelle_Timeline.csv

### Austrian legislative election 2019
https://www.bmi.gv.at/412/Nationalratswahlen/Nationalratswahl_2019/files/endgultiges_Gesamtergebnis_NRW19_16102019.xlsx

### Air quality in Graz (PM10)
http://app.luis.steiermark.at/luft2/suche.php?station1=172&station2=&komponente1=114&station3=&station4=&komponente2=&mittelwert=1

### Individual household electric power consumption
http://archive.ics.uci.edu/ml/datasets/Individual+household+electric+power+consumption
http://archive.ics.uci.edu/ml/machine-learning-databases/00235/household_power_consumption.zip
